Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-12-07 _tex/index.tex
Built site for gh-pages

However, since the discriminal processes of the stimuli are not directly
observable, the theory introduces the \emph{law of comparative
judgment}. This law posits that in pairwise comparisons, a judge
perceives the stimulus positioned further along the trait continuum as
having a higher level of that trait. This emphasizes that the relative
distance between stimuli, rather than their absolute positions on the
continuum, determines the outcome of the pairwise comparison. Indeed,
the theory assumes that the observed dichotomous outcome arises from the
difference between the underlying discriminal processes of the stimuli,
referred to as the \emph{discriminal difference}. Since the individual
discriminal processes follow a Normal distribution, the discriminal
difference also follows a Normal distribution \citep{Andrich_1978}, with
its mode (mean) representing the relative separation between the
stimuli, while its dispersion reflects the variability in that relative
separation.

Figure~\ref{fig-discriminal_difference} shows the distribution of the
discriminal difference for the texts depicted in
Figure~\ref{fig-discriminal_process}. The figure illustrates that the
judge perceives Text B as having significantly higher quality than Text
A, as indicated by their discriminal difference \((S_{B} - S_{A} > 0)\)
and the shaded gray area under the curve \(P(B > A)\). Therefore, the
dichotomous outcome of this comparison would likely favor Text B over
Text A.

@@

The previous section outlines the general form of Thurstone's theory,
which applies to a CJ design where a single judge evaluates multiple
stimuli. For the practical application of the theory, Thurstone
developed four additional cases derived from this general form, where
each successive case incorporates additional simplifying assumptions.
Case I represents the general form of the theory. Case II extends this
by allowing multiple judges to make comparisons rather than restricting
the comparisons to a single judge. Case III introduces the assumption of
zero correlation between stimuli. Case IV builds on this by assuming
that the stimuli have similar dispersions. Finally, Case V replaces this
assumption with the condition that the stimuli have equal discriminal
dispersions. Table~\ref{tbl-thurstone_cases} summarizes these cases and
their assumptions. For a detailed discussion of this progression, refer
to \citet{Thurstone_1927b} and \citet[pp.~248-253]{Bramley_2008}.

@@

Despite its reliance on the largest number of simplifying assumptions
\citetext{\citealp[pp.~253]{Bramley_2008}; \citealp[pp.~677]{Kelly_et_al_2022}},
Case V remains the most widely used case in the CJ literature. This
popularity stems mainly from its simplified statistical representation
in the Bradley-Terry-Luce (BTL) model
\citep{Bradley_et_al_1952, Luce_1959}. The BTL model mirrors the
assumptions of Case V, with one key difference: while Case V assumes a
Normal distribution for the discriminal processes of the stimuli, the
BTL model uses the more mathematically tractable Logistic distribution
\citep[pp.~254]{Andrich_1978, Bramley_2008} (see
Table~\ref{tbl-thurstone_cases}). This substitution has little impact on
the model's estimation or interpretation, as the Normal and Logistic
distributions share similar statistical properties, differing only by a
scaling factor of approximately \(1.7\)
\citep[pp.~16]{vanderLinden_et_al_2017_I} (see
Figure~\ref{fig-logistic_vs_normal}).

@@

Nevertheless, Thurstone originally developed Case V to provide a
``rather coarse scaling'' of traits \citep[pp.~269]{Thurstone_1927b},
prioritizing statistical simplicity over precision in trait measurement
\citep[pp.~677]{Kelly_et_al_2022}. Moreover, he warned against its
untested application, stating that its use ``should not be made without
(an) experimental test'' \citep[pp.~270]{Thurstone_1927b}, acknowledging
that some assumptions might be problematic when complex traits or
heterogeneous stimuli are involved \citep[pp.~376]{Thurstone_1927a}.
Consequently, given that modern CJ applications frequently involve such
traits and stimuli, two main assumptions of Case V and, by extension, of
the BTL model may not consistently hold in theory or practice: the
assumption of equal dispersion and zero correlation between stimuli.

\subsubsection{The assumption of equal dispersions between
stimuli}\label{the-assumption-of-equal-dispersions-between-stimuli}

The discriminal dispersions of stimuli are crucial in determining the
dichotomous outcomes of pairwise comparisons. Discrepancies in these
dispersions shape the distribution of the discriminal difference,
directly influencing the comparison outcome. For instance,
Figure~\ref{fig-dispersion} illustrates how more uncertainty in the
trait perception of a text over another broadens the distribution of
their discriminal difference. This broadening affects the area under the
curve, \(P(B>A)\), determining the comparison outcome. Additionally, the
figure shows that when there are no discrepancies in the texts'
discriminal dispersions, meaning the \emph{texts have equal dispersions}
\((\sigma_{B}-\sigma_{A}=0)\), the likelihood that the discriminal
difference distinctly favors one text over another increases.

@@

However, Thurstone contended that this assumption may not hold when
researchers assess complex traits or heterogeneous stimuli
\citep[pp.~376]{Thurstone_1927a}, as these traits and stimuli can
introduce judgment discrepancies due to their unique features. Indeed,
evidence of such violation may already exist in the CJ literature as
misfit statistics. These statistics measure the judgment discrepancies
associated with a given stimulus
\citetext{\citealp[pp.~12]{Pollitt_2004}; \citealp[pp.~20]{Goossens_et_al_2018}}.
For instance, labeling texts as ``misfits'' indicates that comparisons
involving these texts result in more judgment discrepancies than
comparisons involving other texts
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}.
This finding implies that the discriminal differences associated with
``misfits'' texts usually display a broader dispersion, suggesting that
their discriminal processes also exhibit more variation than other
texts. Notably, this reasoning also applies to ``misfit'' judges, whose
evaluations reflect substantial deviations from the shared consensus due
to the unique characteristics of the stimuli or the judges themselves.

\subsubsection{The assumption of zero correlation between
stimuli}\label{the-assumption-of-zero-correlation-between-stimuli}

Similar to the discriminal dispersions, the correlation between the
discriminal processes is also crucial in determining the dichotomous
outcomes of pairwise comparisons. This correlation shapes the
distribution of the discriminal difference, directly influencing the
comparison outcome. For instance, Figure~\ref{fig-correlation}
illustrates how the stronger dependence of the trait perception in one
text over another narrows the distribution of their discriminal
difference. This narrowing impacts the area under the curve, \(P(B>A)\),
determining the comparison outcome. Additionally, the figure shows that
when the trait perception in one text is independent of another, meaning
the \emph{texts have zero correlation}, the likelihood that the
discriminal difference distinctly favors one text over another is lower
compared to when the texts are highly correlated.

@@

In the first scenario, the intricate aspects of multidimensional,
complex traits may introduce dependencies between the stimuli due to
certain judges' biases that resist cancellation. Research on text
quality suggests that when judges evaluate these traits, they often rely
on various intricate characteristics of the stimuli to form their
judgments
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}. These
additional relevant characteristics, which are unlikely to be equally
weighted or opposing, can unevenly influence judges' perceptions,
creating biases in their judgments and, ultimately, introducing
dependencies between stimuli
\citep[pp.~346]{vanderLinden_et_al_2017_II}. For example, this could
occur when a judge assessing the argumentative quality of a text places
more weight on grammatical accuracy than other judges, ultimately
favoring texts with fewer errors but weaker arguments. While direct
evidence for this specific scenario is lacking, studies such as
\citet{Pollitt_et_al_2003} demonstrate the presence of such biases,
supporting the idea that the factors influencing pairwise comparisons
may not always cancel out.

In the second scenario, the shared context or inherent connections
created by additional hierarchical structures may further introduce
dependencies between stimuli, a statistical phenomenon commonly known as
clustering \citep{Everitt_et_al_2010}. Although the CJ literature
acknowledges the presence of such hierarchical structures, the
statistical handling of this extra source of dependency between stimuli
has been inadequate. For example, when CJ data includes multiple samples
of stimuli from the same individuals, researchers often rely on
(average) estimated BTL scores to conduct subsequent analyses and tests
at the individual hierarchical level
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
This approach, however, has the significant limitation of ignoring the
uncertainty associated with the BTL scores, which generates additional
statistical and measurement issues, as discussed in section
Section~\ref{sec-theory-issue2}.

@@

Building on the previous section, it is evident that, besides its
inconveniences, the BTL model commonly functions as the trait's
measurement model in CJ assessments. A measurement model specifies how
manifest variables contribute to the estimation of latent variables
\citep{Everitt_et_al_2010}. For example, when evaluating text quality,
researchers use the BTL model to process the dichotomous outcomes
resulting from the pairwise comparisons (the manifest variables) to
estimate scores that reflect the underlying quality level of texts (the
latent variable)
\citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.

>>>>

However, since the discriminal processes of the stimuli are not directly
observable, the theory introduces the \emph{law of comparative
judgment}. This law states that in pairwise comparisons, a judge
perceives the stimulus positioned further along the trait continuum as
having a higher level of that trait. This law emphasizes that the
relative distance between stimuli, rather than their absolute positions
on the continuum, determines the outcome of the pairwise comparison.
Indeed, the theory assumes that the observed dichotomous outcome arises
from the difference between the underlying discriminal processes of the
stimuli, referred to as the \emph{discriminal difference}. Since the
individual discriminal processes follow a Normal distribution, the
discriminal difference also follows a Normal distribution
\citep{Andrich_1978}. In this distribution, the mode (mean) represents
the relative separation between the stimuli and the dispersion captures
the variability of that relative separation.

Figure~\ref{fig-discriminal_difference} illustrates the distribution of
the discriminal difference for the texts presented in
Figure~\ref{fig-discriminal_process}. The figure indicates that the
judge perceives Text B as having significantly higher quality than Text
A. This conclusion is evident from the positive difference in the modal
discriminal processes of the texts \((S_{B} - S_{A} > 0)\) and the area
under the curve where the discriminal difference distinctly favors Text
B over Text A, denoted as \(P(B > A)\) (shaded gray area). As a result,
the dichotomous outcome of this comparison is more likely to favor Text
B over Text A.

@@

The previous section outlines the general form of Thurstone's theory,
which applies to a CJ design where a single judge evaluates multiple
stimuli. For the practical application of the theory, Thurstone
developed four additional cases derived from this general form, where
each successive case incorporates additional simplifying assumptions.
Case I represents the general form of the theory. Case II extends this
by allowing multiple judges to make comparisons rather than restricting
the comparisons to a single judge. Case III introduces the assumption of
zero correlation between stimuli. Case IV builds on this by assuming the
stimuli have similar dispersions. Finally, Case V replaces this
assumption with the condition that the stimuli have equal discriminal
dispersions. Table~\ref{tbl-thurstone_cases} summarizes these cases and
their assumptions. For a detailed discussion of this progression, refer
to \citet{Thurstone_1927b} and \citet[pp.~248-253]{Bramley_2008}.

@@

Despite its reliance on the largest number of simplifying assumptions
\citetext{\citealp[pp.~253]{Bramley_2008}; \citealp[pp.~677]{Kelly_et_al_2022}},
Case V remains the most widely used case in the CJ literature. This
popularity stems mainly from its simplified statistical representation
in the Bradley-Terry-Luce (BTL) model
\citep{Bradley_et_al_1952, Luce_1959}. The BTL model mirrors the
assumptions of Case V, with one key difference: while Case V assumes a
Normal distribution for the stimuli's discriminal processes, the BTL
model uses the more mathematically tractable Logistic distribution
\citep[pp.~254]{Andrich_1978, Bramley_2008} (see
Table~\ref{tbl-thurstone_cases}). This substitution has little impact on
the model's estimation or interpretation, as the Normal and Logistic
distributions share similar statistical properties, differing only by a
scaling factor of approximately \(1.7\)
\citep[pp.~16]{vanderLinden_et_al_2017_I} (see
Figure~\ref{fig-logistic_vs_normal}).

@@

Nevertheless, Thurstone originally developed Case V to provide a
``rather coarse scaling'' of traits \citep[pp.~269]{Thurstone_1927b},
prioritizing statistical simplicity over precision in trait measurement
\citep[pp.~677]{Kelly_et_al_2022}. He explicitly warned against its
untested application, stating that its use ``should not be made without
(an) experimental test'' \citep[pp.~270]{Thurstone_1927b}, acknowledging
that some assumptions might be problematic when complex traits or
heterogeneous stimuli are involved \citep[pp.~376]{Thurstone_1927a}.
Consequently, given that modern CJ applications frequently involve such
traits and stimuli, two main assumptions of Case V and, by extension, of
the BTL model may not consistently hold in theory or practice: the
assumption of equal dispersion and zero correlation between stimuli.

\subsubsection{The assumption of equal dispersions between
stimuli}\label{sec-theory-issue1a}

The discriminal dispersions of stimuli are crucial in determining the
dichotomous outcomes of pairwise comparisons. Discrepancies in these
dispersions shape the distribution of the discriminal difference,
directly influencing the comparison outcome. Figure~\ref{fig-dispersion}
illustrates how more uncertainty in the trait perception of one text
relative to another broadens the distribution of their discriminal
difference. This broadening affects the area under the curve where the
discriminal difference distinctly favors one text over the other,
\(P(B > A)\). Additionally, the figure shows that when the discriminal
dispersions of the texts are equal \((\sigma_{B}-\sigma_{A}=0)\), the
discriminal difference is more likely to favor one text over the other
than when their dispersions differ (shaded gray area).

@@

However, Thurstone contended that the assumption of equal dispersions
may not hold when researchers assess complex traits or heterogeneous
stimuli \citep[pp.~376]{Thurstone_1927a}, as these traits and stimuli
can introduce judgment discrepancies due to their unique features
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}.
Indeed, evidence of such violation may already exist in the CJ
literature as misfit statistics. These statistics measure the judgment
discrepancies associated with a given stimulus
\citetext{\citealp[pp.~12]{Pollitt_2004}; \citealp[pp.~20]{Goossens_et_al_2018}}.
For instance, labeling texts as ``misfits'' indicates that comparisons
involving these texts result in more judgment discrepancies than others
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}.
This finding implies that discriminal differences associated with
``misfits'' texts usually display a broader dispersion, suggesting that
their discriminal processes also exhibit more variation than other
texts. Notably, this reasoning also applies to ``misfit'' judges, whose
evaluations reflect substantial deviations from the shared consensus due
to the unique characteristics of the stimuli or the judges themselves.
Moreover, the presence of these ``misfit'' judges and their deviations
can introduce additional statistical and measurement issues, as
discussed in Section~\ref{sec-theory-issue1b}.

Therefore, assuming equal dispersions between stimuli, despite its
violation, can cause Case V (and the BTL model) to inflate the
reliability of the comparison outcome. Moreover, ignoring the difference
in dispersions can lead to the neglect of critical differences in the
reliability of the trait across stimuli, resulting in inaccurate
conclusions about the trait's estimates \citep[pp.~370]{McElreath_2020}.
Furthermore, if researchers acknowledge that misfit statistics help
identify these critical differences, the usual practice in the CJ
literature of excluding stimuli based on these statistics
\citep{Pollitt_2012b, vanDaal_et_al_2017, Goossens_et_al_2018} risks
discarding valuable information and introducing bias into the trait's
estimates \citep[chap.~12]{Zimmerman_1994, McElreath_2020}. The
direction and magnitude of these biases are unpredictable, as they
depend on the specific stimuli researchers exclude from the analysis.
Together, these oversights undermine the reliability of the trait and
ultimately compromise its validity \citep[pp.~2]{Perron_et_al_2015}.

\subsubsection{The assumption of zero correlation between
stimuli}\label{sec-theory-issue1b}

Similar to the discriminal dispersions, the correlation between
discriminal processes plays a crucial role in determining the
dichotomous outcomes of pairwise comparisons. This correlation also
shapes the distribution of the discriminal difference, directly
influencing the comparison outcome. Figure~\ref{fig-correlation}
illustrates how the dependence in trait perception between two texts
narrows the distribution of their discriminal difference. This narrowing
affects the area under the curve where the discriminal difference
distinctly favors one text over the other, \(P(B>A)\). Moreover, the
figure shows that when two texts are independent or uncorrelated
\((\rho=0)\), their discriminal difference is less likely to favor one
text over the other than when the texts are highly correlated (shaded
gray area).

@@

In the first scenario, the intricate aspects of multidimensional,
complex traits may introduce dependencies between the stimuli due to
certain judges' biases that resist cancellation. Research on text
quality suggests that when judges evaluate these traits, they often rely
on various intricate characteristics of the stimuli to form their
judgments
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}. These
additional relevant characteristics, which are unlikely to be equally
weighted or opposing, can unevenly influence judges' perceptions,
creating biases in their judgments and, ultimately, introducing
dependencies between stimuli
\citep[pp.~346]{vanderLinden_et_al_2017_II}. For example, this could
occur when a judge assessing the argumentative quality of a text places
more weight on its grammatical accuracy than other judges, ultimately
favoring texts with fewer errors but weaker arguments. While direct
evidence for this specific scenario is lacking, studies such as
\citet{Pollitt_et_al_2003} demonstrate the presence of such biases,
supporting the idea that the factors influencing pairwise comparisons
may not always cancel out.

In the second scenario, the shared context or inherent connections
created by additional hierarchical structures may further introduce
dependencies between stimuli, a statistical phenomenon commonly known as
clustering \citep{Everitt_et_al_2010}. Although the CJ literature
acknowledges the presence of such hierarchical structures, the
statistical handling of this extra source of dependency between stimuli
has been inadequate. For example, when CJ data includes multiple samples
of stimuli from the same individuals, researchers often rely on
(average) estimated BTL scores to conduct subsequent analyses and tests
at the individual hierarchical level
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
This approach, however, can introduce additional statistical and
measurement issues, as discussed in Section~\ref{sec-theory-issue2}.

In any case, the psychometric and statistical literature emphasizes the
need to address factors that create dependencies between stimuli, as
failing to do so can affect the reliability of the comparison outcomes
and lead to inaccurate conclusions about the trait's estimates. For
instance, researchers who overlook additional relevant traits, such as
judges' biases, can cause dimensional mismatches in the statistical
model used for analysis. This mismatch can artificially inflate the
reliability of the trait \citep[pp.~341]{Hoyle_et_al_2023} or, even
worse, introduce bias into the trait's estimates \citep{Ackerman_1989}.
Furthermore, as discussed in Section~\ref{sec-theory-issue1a},
researchers who exclude judges based on misfit statistics can risk
discarding valuable information, further biasing the trait's estimates
\citep[chap.~12]{Zimmerman_1994, McElreath_2020}. Lastly, researchers
who fail to account for hierarchical (grouping) structures can reduce
the precision of model parameter estimates, which may amplify the
overestimation of reliability \citep[pp.~482]{Hoyle_et_al_2023}.
Together, these issues undermine the reliability of the trait and
compromise the validity of its estimates
\citep[pp.~2]{Perron_et_al_2015}.

@@

Building on the previous section, it is clear that, despite its
limitations, the BTL model is commonly used as the measurement model in
CJ assessments. A measurement model specifies how manifest variables
contribute to the estimation of latent variables
\citep{Everitt_et_al_2010}. For example, when evaluating text quality,
researchers use the BTL model to process the dichotomous outcomes
resulting from the pairwise comparisons (the manifest variables) to
estimate scores that reflect the underlying quality level of the texts
(the latent variable)
\citep{Laming_2004, Pollitt_2012b, Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}.
